---
title: "HW4"
author: "Howard"
date: "2025-03-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Problem 5

### Create the data
```{r, warning=F}
# Data for Problem 5 - Painted Turtles

# Female turtles data
female <- data.frame(
  Length = c(98, 103, 103, 105, 109, 123, 123, 133, 133, 133, 134, 136, 138, 138, 141, 147, 149, 153, 155, 155, 158, 159, 162, 177),
  Width  = c(81, 84, 86, 86, 88, 92, 95, 99, 102, 102, 100, 102, 98, 99, 105, 108, 107, 107, 115, 117, 115, 118, 124, 132),
  Height = c(38, 38, 42, 42, 44, 50, 46, 51, 51, 51, 48, 49, 51, 51, 53, 57, 55, 56, 63, 60, 62, 63, 61, 67)
)

# Male turtles data
male <- data.frame(
  Length = c(93, 94, 96, 101, 102, 103, 104, 106, 107, 112, 113, 114, 116, 117, 117, 119, 120, 120, 121, 125, 127, 128, 131, 135),
  Width  = c(74, 78, 80, 84, 85, 81, 83, 83, 82, 89, 88, 86, 90, 90, 91, 93, 89, 93, 95, 93, 96, 95, 95, 106),
  Height = c(37, 35, 35, 39, 38, 37, 39, 39, 38, 40, 40, 40, 43, 41, 41, 41, 40, 44, 42, 45, 45, 45, 46, 47)
)

# Combine the female and male data into one data frame with a grouping variable
turtles <- rbind(
  data.frame(Group = "Female", female),
  data.frame(Group = "Male", male)
)

# View the combined dataset
head(turtles)

```

### Log-transform the measurements
```{r}
turtles$logLength <- log(turtles$Length)
turtles$logWidth  <- log(turtles$Width)
turtles$logHeight <- log(turtles$Height)
```

### Look at group means on the log scale
```{r}
group_means <- aggregate(cbind(logLength, logWidth, logHeight) ~ Group, data = turtles, FUN = mean)
print(group_means)
```

## **\(H_0: \mu_F=\mu_M \ versus \ H_1: \mu_F \neq \mu_M \)**

### Fit a MANOVA model using the log-transformed variables
```{r}
fit <- manova(cbind(logLength, logWidth, logHeight) ~ Group, data = turtles)

# Summary using Wilks' lambda test
summary_fit <- summary(fit, test = "Wilks")
print(summary_fit)

```
### Conclution
Based on the extremely small p‑value (on the order of \(10^{-10}\)), we would **reject** the null hypothesis \(H_0\) that female and male turtles have the same mean (log‑transformed) carapace measurements. In other words, **there is a statistically significant difference** between the mean vectors of female and male turtles at the 5% significance level.

################################
## Problem 6                   #
################################
### Data setup
```{r}
# 1. Create vectors for species, nutrient, and the two reflectance variables
species <- factor(c("SS", "JL", "LP", "SS", "JL", "LP"))
nutrient <- factor(c("+", "+", "+", "-", "-", "-"))
x560 <- c(10.35, 13.41, 7.78, 10.40, 17.78, 10.40)
x720 <- c(25.93, 38.63, 25.15, 24.25, 41.45, 29.20)

# 2. Combine into a single data frame
seedlings <- data.frame(x560, x720,species, nutrient)
print(seedlings)
```

### (a)Separate MANOVA TEST

#### MANOVA for the Effect of Species

- Test the effect of species (SS, JL, LP), ignoring nutrient.

```{r}
# MANOVA for species only
fit_species <- manova(cbind(x560, x720) ~ species, data = seedlings)
summary_species <- summary(fit_species, test = "Wilks")
summary_species

```
- **Interpretation**: The p‑value is < 0.05, we reject the null hypothesis that all **three species** have the same mean vector \((\mu_{560},\mu_{720})\)

#### MANOVA for the Effect of Nutrient

- Test The effect of nutrient (+, −), ignoring species.

```{r}
# MANOVA for nutrient only
fit_nutrient <- manova(cbind(x560, x720) ~ nutrient, data = seedlings)
summary_nutrient <- summary(fit_nutrient, test = "Wilks")
summary_nutrient

```
- **Interpretation**: The p‑value is > 0.05, we do reject the null hypothesis that  **the two nutrient levels** have the same mean vector \((\mu_{560},\mu_{720})\)

### (b) Two-Way ANOVA for 560CM and 720CM Separately

#### ANOVA for 560CM
```{r}
# Two-way ANOVA for 560CM (no interaction due to no replication)
fit_aov_560 <- aov(x560 ~ species + nutrient, data = seedlings)
summary(fit_aov_560)
```
#### **Interpretation**: 
- *Species (p = 0.0905)*: At the conventional 5% significance level, we do not reject the null hypothesis that all three species (SS, JL, LP) have the same 560 nm mean reflectance.

- *Nutrient (p = 0.2023)*: There is no evidence at the 5% level to conclude that the two nutrient levels (+, −) differ in their 560 nm mean reflectance.

#### ANOVA for 720CM
```{r}
# Two-way ANOVA for 720CM (no interaction due to no replication)
fit_aov_720 <- aov(x720 ~ species + nutrient, data = seedlings)
summary(fit_aov_720)
```
#### **Interpretation**: 
- *Species (p = 0.0335)*: At the 5% significance level, we reject the null hypothesis that the three species (SS, JL, LP) have the same mean reflectance at 720 nm. In other words, there is evidence of a significant difference among species in their 720 nm reflectance means.

- *Nutrient (p = 0.4252)*:  There is no evidence at the 5% level to conclude that the two nutrient levels (+, −) differ in their 720 nm mean reflectance.


### Comparing MANOVA vs. Separate ANOVAs
- MANOVA results tell you if the factor(s) has/have a joint effect on the vector \(x_{560},x_{720}\)


- Separate ANOVAs on `x560` and `x720` test each response individually.



